---
title: "Calibrating the natural-history model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating the natural-history model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Every projection in `mbiedesign` starts from a **calibrated natural-history
model** — a continuous-time multistate model whose transition rates are fit by
maximum likelihood to age- and stage-specific cancer incidence. This vignette
shows the full path from raw incidence data to a fitted model.

For convenience, `fit_natural_history()` and `load_fitted_model()` do the data
handling for you. Here we open that up so you can see exactly what happens —
and adapt it to your own data.

```{r setup}
library(mbiedesign)
```

## 1. The incidence data

The package includes SEER-style lung cancer incidence, one row per age band and
stage:

```{r read}
xlsx <- system.file("extdata", "lung_incidence.xlsx", package = "mbiedesign")
inc  <- openxlsx::read.xlsx(xlsx)
head(inc)
```

SEER\*Stat gives incidence by 5-year age band, with case counts, populations,
and rates. The model needs five columns (one row per age band and stage):

- **`midage`** — midpoint of the age band, in years (e.g. 62.5 for 60–64),
- **`Count`** — number of incident cancers in that age band and stage,
- **`Pop`** — population (denominator),
- **`years`** — person-time multiplier used to form the denominator
  (person-years = `Pop * years`),
- **`stage`** — stage at diagnosis, in two levels: **`"Early"`** for early-stage
  rows and **`"Advanced"`** for late-stage rows.

**The early vs advanced split is your modeling choice.** Label early-stage rows
`"Early"` and late-stage rows `"Advanced"`; *which* stages count as early is up
to you. In this paper we used AJCC7 stage I–II as
early and III–IV as advanced, but you might instead use SEER Summary Stage,
AJCC6 or AJCC9, or a different cut point (for example only stage I as early).
How you handle unstaged or missing-stage records is likewise your choice — in
this paper we distributed them across early and advanced in proportion to the
observed (known-stage) distribution. Any extra columns (here `age`, `Rate`) are
ignored. To calibrate a **different cancer site**, export the same age-by-stage
incidence from SEER\*Stat and supply it via `the_data` (see the last section).

## 2. Preparing the data

Two derived quantities and one filter turn the raw counts into what the
likelihood uses:

```{r prep}
risk_multiplier <- 3.1   # incidence inflation factor (see below)

inc$PY    <- inc$Pop * inc$years                      # person-years
inc$rate  <- risk_multiplier * inc$Count / inc$PY * 1e5   # rate per 100,000
inc$Count <- inc$Count * risk_multiplier              # inflate the counts too
inc <- inc[inc$midage > 0 & inc$midage < 80, ]        # restrict to ages 0-80

head(inc[c("midage", "stage", "Count", "PY", "rate")])
```

The **risk multiplier** is the one modeling choice worth dwelling on. General-
population SEER incidence understates the risk in a *screening-trial* population,
which is enriched for risk factors (in the NLST, heavy smoking history). The
multiplier of `3.1` inflates incidence to reflect that NLST-eligible high-risk
cohort. It is a parameter you can change — a different trial population, or a
different cancer, will warrant a different value.

This is exactly what `fit_natural_history()` does internally via its `cancer`
and `risk_multiplier` arguments; the block above simply makes it explicit.

## 3. Fitting the model

With the prepared data, calibration is a single call. You specify the two target
sojourn times — the overall mean sojourn time (`OMST`) and the late-stage mean
sojourn time (`LMST`) — and the fitter finds transition rates consistent with
them that best reproduce the observed incidence:

```{r fit, eval = FALSE}
# Equivalent one-liner (loads + prepares the provided data for you):
fit <- fit_natural_history(OMST = 4.0, LMST = 1.35)

# ...or fit the data we prepared above explicitly:
fit <- fit_natural_history(OMST = 4.0, LMST = 1.35, the_data = inc)
```

The optimizer arguments are optional with sensible defaults:

- **`k`** (default 16) — number of states in the model,
- **`num_seeds`** (default 1) — random optimizer restarts; the best fit is kept,
- **`mean1`** (default 3) — mean of the distribution used to draw starting values,
- **`risk_multiplier`** (default 3.1) — the inflation factor from Section 2,
- **`the_data`** — supply your own incidence frame to override the provided `cancer`.

Fitting runs a maximum-likelihood optimization, so it takes a little time; the
chunk above is not evaluated when this vignette builds. The package provides the
result of exactly this calibration.

## 4. Inspecting the fitted model

Load the provided fit and look at what calibration produced:

```{r loadfit}
fit <- load_fitted_model("lung")$base_fit

# The fitted transition-rate matrix (the object projections consume)
dim(fit$rate.matrix)

# Sojourn-time summaries, including the early-stage mean sojourn time (EMST)
str(fit$summary_out)
```

The `rate.matrix` is the hand-off point: pass it to
`get_trial_results_by_design()` or `get_summary_outcomes_by_design()` to project trial outcomes
(see `vignette("mbiedesign")`).

## Fitting your own data or another cancer site

The provided model is lung, but the natural-history model calibrates to **any
cancer site**. Prepare an incidence data frame with the five columns from
Section 1 — `midage`, `Count`, `Pop`, `years`, `stage` (`"Early"` vs
`"Advanced"`) — typically a SEER\*Stat age-by-stage incidence export for your
site, then pass it directly:

```{r own, eval = FALSE}
# OMST and LMST are your choice; use sojourn times reasonable for your cancer,
# not the lung values below.
fit <- fit_natural_history(OMST = 4.0, LMST = 1.35, the_data = my_incidence)
```

The `OMST` and `LMST` targets are modeling inputs *you* choose — reasonable
values depend on the cancer and should come from prior estimates or the
literature for your site; the 4.0 / 1.35 used here are lung-specific. No
`risk_multiplier` is applied to data you supply yourself — scale it as
appropriate for your target population before fitting.
