---
title: "Getting started with mbiedesign"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mbiedesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`mbiedesign` helps you design and size cancer screening trials that use
**late-stage cancer incidence** as the endpoint. You calibrate a natural-history
model, then use one function — `get_summary_outcomes_by_design()` — to project
the trial outcomes and required per-arm sample size for the traditional and
intended-effect (IE) designs.

```{r setup}
library(mbiedesign)
```

## 1. Choose a calibrated model

A calibrated model is defined by two mean sojourn times you choose — the
**overall mean sojourn time (OMST)** and the **late-stage mean sojourn time
(LMST)**; the **early-stage mean sojourn time (EMST)** is derived by the fit. A
lung cancer model, calibrated to SEER age- and stage-specific incidence, is
available at the base case and across the sensitivity analyses, so you can
select one directly:

```{r load}
model <- load_fitted_model("lung", OMST = 4.0, LMST = 1.35)

model$summary_out$OMST   # overall mean sojourn time     (chosen)  = 4.00 years
model$summary_out$LMST   # late-stage mean sojourn time  (chosen)  = 1.35 years
model$summary_out$EMST   # early-stage mean sojourn time (derived) = 3.01 years
```

For sojourn times that are not precalibrated — or a different cancer site —
calibrate a new model with `fit_natural_history()` (see `vignette("calibration")`):

```{r fit-call, eval = FALSE}
model <- fit_natural_history(OMST = 3.5, LMST = 1.2)
```

## 2. Summary outcomes and sample size by design

`get_summary_outcomes_by_design()` is the main function. For a given screening
schedule and test, it returns one row per design — traditional and IE —
with the late-stage event rates, the required per-arm sample size, and the
**relative efficiency** (N required by the traditional design divided by N
required by that design):

```{r summary}
summary_data <- get_summary_outcomes_by_design(
  rate_matrix = model$rate.matrix,
  start_age = 62, numscreens = 3, screen_int = 1, num_followup_intervals = 4.5,
  sens_e = 0.35, sens_l = 0.82, specificity = 0.855,   # test performance
  allocation_ratio = 1,                                 # balanced arms
  power = 0.90, alpha = 0.025                           # target power (= 1 - beta), one-sided level
)

summary_data[c("design", "fraction", "n_per_arm", "relative_efficiency")]
```

A relative efficiency above 1 means that design reaches the same power with
fewer participants than the traditional design. Pass `design = "IE"` to return a
single design. Each row also carries the control- and screen-arm event rates
(`p_c`, `p_s`), the risk ratio and difference (`rr`, `rd`), and the subgroup
fraction the design conditions on (`fraction` — the ever-positive fraction for
IE).

## 3. The full projection (optional)

For the early-, late-, and overall-stage outcomes over the course of the trial,
use `get_trial_results_by_design()`, then `stage_shift_by_design()` to extract a
table for one design:

```{r project}
trial_results <- get_trial_results_by_design(
  rate_matrix = model$rate.matrix,
  start_age = 62, numscreens = 3, screen_int = 1, num_followup_intervals = 4.5,
  sens_e = 0.35, sens_l = 0.82, specificity = 0.855,
  n_control = 26722, n_screen = 26722
)

head(stage_shift_by_design(trial_results, design = "standard",
                           yearly_or_cumulative = "cumulative"))
```

## Where to go next

- `vignette("calibration")` — how the natural-history model is fit to incidence
  data, and how to calibrate to your own.
- `vignette("manuscript-analysis")` — the full late-stage lung cancer results,
  comparing the traditional and intended-effect designs, with sensitivity
  analyses.
